Project Report\\\textsc{DEVS Ex Machina}

### 2015-04-25 manual/projectrapport.tex
Update Design.dia and png, smallish updates in tex.

\documentclass[8pt,a4paper]{report}
\usepackage[latin1]{inputenc}
\usepackage{xcolor}
\usepackage{textcomp}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{listings}
\usepackage{amssymb}
\usepackage[T1]{fontenc}
\usepackage[scaled]{beramono}
\newcommand\Small{\fontsize{9}{9.2}\selectfont}
\newcommand*\LSTfont{\Small\ttfamily\SetTracking{encoding=*}{-60}\lsstyle}
\usepackage{url}
  \lstset{basicstyle=\ttfamily\tiny\color{blue}}
\usepackage{graphicx}
\usepackage[]{algorithm2e}

\title{Project Rapport\\\textsc{DEVS Ex Machina}}
\author{Ben Cardoen\\Pieter Lauwers\\Stijn Manhaeve\\Tim Tuijn\\Matthijs Van Os}
\date{}
\begin{document}
\maketitle
\tableofcontents{}

@@

\section{Network -- Kernel communication}
PyPDEVS uses MPI to exchange messages between kernels, whereas we could use shared memory to communicate between kernels (each running on 1 or more threads). We created a network architecture mimicking the approach used by PyPDEVS, without sacrificing the speed and ordering advantages of shared memory. The Network class consists of a single locked receiving port, and has for each existing kernel an outgoing queue. Receiving a message locks the network for the time it needs to place it on an integer indexed queue (very fast). If a kernel wants to receive messages it can do so if/when it is ready to do so, this only locks if at the same time a message is being received by the network.\\ Several kernels can pull messages lockfree in parallel from the network (since the output queues are not shared). Early benchmarks made a throughput of 5e5 msg/s repeatable, with 4 kernels on modest hardware. The problems of (unintentional) out of order receipt and potential loss of messages are a non-issue in this context, since we use FIFO queues and loss of a message would implicate severe system failure (in which case message loss is the lesser problem).\\
Another advantage of hiding the message exchange behind a shared Network class instance is that, should this be required in the future, the kernel could run on different machines without requiring much redesigning of the system. An RPC implementation would simply have to implement the Network class on both sides.
\includegraphics[scale=0.2]{network.png}

@@

\chapter{Obstacles}
\section{Virtualization}
Virtualization proved a significant hurdle by exposing potential faults. Standard compliant code was found to be leaking memory by not calling container member destructors but only on a VM with Linux as Guest and Windows as Host. This was never reproducible in another setting.
\section{Threading}
G++'s std::string implementation appears to be still Copy On Write (despite 21.4.1 p6 from the C++ standard) \cite{cpp}, triggering an alert by threadsanitizer for a possible race whenever a string was supposedly copied but internally still stored as a reference to the same object. Explicitly invoking the string constructor with the char[] proved to be a safe workaround here. Note that this problem is a known issue \cite{cow}
\section{Toolchains}
To develop we used the required toolchains:\\ g++/clang++ as compiler, Eclipse+CDT as IDE, git for dvcs, jenkins as ci, cmake as build tool supplemented with some shell and python scripting.
\section{Logger}
The recommended logger (g3log), while powerful and fast, proved a dead end since it does not compile on Cygwin. 
Stijn therefore wrote his own implementation. %TODO Stijn
\chapter{Performance Analysis}
Whenever we made big design/architecture decisions, we first wrote some prototyping testcode to prove the viability (e.g. Network, Tracing, Logger).
\section{First comparison}
We use perf to get statistically sound benchmarks.
\subsection{Classic}
An endtime of 3.6e6 was set, and both examples run against each other.
\lstinputlisting[]{pycla.txt}
The following shows a recording of the C++ implementation of the same example.\\
\lstinputlisting[]{cpcla.txt}
\subsubsection{Conclusion}
The C++ port is approximately 12.906 times faster.

>>>>

\documentclass[8pt,a4paper]{report}
\usepackage[latin1]{inputenc}
\usepackage{xcolor}
\usepackage{textcomp}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{listings}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage[T1]{fontenc}
\usepackage[scaled]{beramono}
\newcommand\Small{\fontsize{9}{9.2}\selectfont}
\newcommand*\LSTfont{\Small\ttfamily\SetTracking{encoding=*}{-60}\lsstyle}
\usepackage{url}
  \lstset{basicstyle=\ttfamily\tiny\color{blue}}
\usepackage{graphicx}
\usepackage[]{algorithm2e}

\title{Project Report\\\textsc{DEVS Ex Machina}}
\author{Ben Cardoen\\Pieter Lauwers\\Stijn Manhaeve\\Tim Tuijn\\Matthijs Van Os}
\date{}
\begin{document}
\maketitle
\tableofcontents{}

@@

\section{Network -- Kernel communication}
PyPDEVS uses MPI to exchange messages between kernels, whereas we could use shared memory to communicate between kernels (each running on 1 or more threads). We created a network architecture mimicking the approach used by PyPDEVS, without sacrificing the speed and ordering advantages of shared memory. The Network class consists of a single locked receiving port, and has for each existing kernel an outgoing queue. Receiving a message locks the network for the time it needs to place it on an integer indexed queue (very fast). If a kernel wants to receive messages it can do so if/when it is ready to do so, this only locks if at the same time a message is being received by the network.\\ Several kernels can pull messages without contention in parallel from the network (since the output queues are not shared). \\This is illustrated in \hyperref[networkdesign]{Network design figure \ref*{networkdesign}}.

\begin{figure}[h!]
	\makebox[\textwidth][c]{\includegraphics[width=\textwidth]{network.png}}
	\caption{Network cooperation diagram}
	\label{networkdesign}
\end{figure} 
Early benchmarks showed a repeatable throughput of 5e5 msg/s, with 4 kernels on modest hardware. The problems of (unintentional) out of order receipt and potential loss of messages are a non-issue in this context, since we use FIFO queues and loss of a message would implicate severe system failure (in which case message loss is the lesser problem).\\
Another advantage of hiding the message exchange behind a shared Network class instance is that, should this be required in the future, the kernel could run on different machines without requiring much redesigning of the system. An RPC implementation would simply have to implement the Network class on both sides.

@@

\chapter{Obstacles}
\section{Virtualization}
Virtualization proved a significant hurdle by exposing potential faults. Standard compliant code was found to be leaking memory by not calling container member destructors but only on a VM with Linux as Guest and Windows as Host. This was never reproducible in another setting.
\section{Threading}
libstdc++'s std::string implementation appears to be still Copy On Write (despite 21.4.1 p6 from the C++ standard) \cite{cpp}, triggering an alert by threadsanitizer for a possible race whenever a string was supposedly copied but internally still stored as a reference to the same object. Explicitly invoking the string constructor with the char[] proved to be a safe workaround here. Note that this problem is a known issue \cite{cow}.
\section{Toolchains}
To develop we used the required toolchains:\\ g++/clang++ as compiler, Eclipse+CDT as IDE, git for dvcs, jenkins as ci, cmake as build tool supplemented with some shell and python scripting.
\section{Logger}
The recommended logger (g3log), while powerful and fast, proved a dead end since it does not compile on Cygwin. 
Stijn therefore wrote his own implementation. %TODO Stijn
\chapter{Performance Analysis}
Whenever we made big design/architecture decisions, we first wrote some prototyping testcode to prove the viability (e.g. Network, Tracing, Logger).
\section{First comparison}
We use perf to get statistically sound benchmarks.
\subsection{Classic}
An endtime of 3.6e6 was set, and both examples run against each other.
\lstinputlisting[]{pycla.txt}
The following shows a recording of the C++ implementation of the same example.\\
\lstinputlisting[]{cpcla.txt}
\subsubsection{Conclusion}
The C++ port is approximately 12.906 times faster.
